---
title: "Predicting Stroke Risk Using Supervised Machine Learning"
author: "Islam Gamal"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    latex_engine: xelatex
    fig_caption: true
    highlight: tango
  html_document:
    toc: true
    theme: flatly
fontsize: 11pt
geometry: margin=1in
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo       = TRUE,
  warning    = FALSE,
  message    = FALSE,
  fig.align  = "center",
  fig.width  = 8,
  fig.height = 5,
  cache      = FALSE
)
```

```{r packages, include=FALSE}
packages <- c("tidyverse", "caret", "randomForest", "pROC",
              "ggplot2", "gridExtra", "corrplot", "e1071",
              "ROSE", "knitr", "kableExtra", "scales")

for (pkg in packages) {
  if (!require(pkg, character.only = TRUE, quietly = TRUE)) {
    install.packages(pkg, repos = "https://cloud.r-project.org")
    library(pkg, character.only = TRUE)
  }
}
```

\newpage

# Introduction

## Background and Motivation

Stroke is one of the leading causes of death and long-term disability worldwide. According to the World Health Organization, stroke accounts for approximately 11% of total deaths globally. What makes this particularly tragic is that a large proportion of strokes are preventable — but only if high-risk individuals are identified before the event occurs.

I chose this dataset and problem because I genuinely find healthcare machine learning compelling, and because the challenge of predicting a rare but life-altering event forces you to think carefully about model design, evaluation strategy, and clinical interpretation. This isn't just an exercise in accuracy — it's about building something that could matter.

## Dataset Description

The dataset I used is the **Stroke Prediction Dataset** from Kaggle, originally compiled for clinical research purposes. It contains **5,110 patient records** with **12 variables**, including demographic information, medical history, and lifestyle factors.

Here is a brief description of each variable:

| Variable | Type | Description |
|---|---|---|
| `id` | Integer | Patient identifier (excluded from modeling) |
| `gender` | Categorical | Male, Female, Other |
| `age` | Numeric | Patient age in years |
| `hypertension` | Binary | Whether patient has hypertension |
| `heart_disease` | Binary | Whether patient has heart disease |
| `ever_married` | Binary | Marital status (Yes/No) |
| `work_type` | Categorical | Employment type |
| `Residence_type` | Categorical | Urban or Rural |
| `avg_glucose_level` | Numeric | Average blood glucose (mg/dL) |
| `bmi` | Numeric | Body Mass Index |
| `smoking_status` | Categorical | Never, Formerly, Smokes |
| `stroke` | Binary | **Target variable**: 1 = Stroke, 0 = No stroke |

## Problem Definition

The goal of this project is to build a binary classification model that predicts whether a patient is likely to experience a stroke, based on the variables above. This is a supervised learning problem with a significant class imbalance challenge — only about 5% of patients in the dataset experienced a stroke.

## Why This Matters

A model like this could serve as a screening tool in general practice, flagging patients who warrant closer monitoring or preventive intervention. Even a modest improvement in early identification could translate into thousands of prevented strokes each year.

\newpage

# Methods and Analysis

## Data Loading

I automated the dataset download to ensure full reproducibility. The dataset is pulled from a public GitHub mirror of the original Kaggle source.

```{r load-data}
data_url <- paste0(
  "https://raw.githubusercontent.com/",
  "Rohit-Gupta-Web3/Stroke-Prediction-Dataset/",
  "main/healthcare-dataset-stroke-data.csv"
)

data_file <- "stroke_data.csv"

if (!file.exists(data_file)) {
  tryCatch({
    download.file(data_url, destfile = data_file, mode = "wb")
    cat("Dataset downloaded successfully.\n")
  }, error = function(e) {
    stop("Download failed. Please check your internet connection.")
  })
} else {
  cat("Dataset loaded from local cache.\n")
}

stroke_raw <- read.csv(data_file, stringsAsFactors = FALSE)
cat("Dimensions:", nrow(stroke_raw), "rows x", ncol(stroke_raw), "columns\n")
```

## Data Cleaning

### Missing Values

```{r missing-values}
# BMI column contains "N/A" strings
stroke_raw$bmi <- as.numeric(stroke_raw$bmi)
cat("Missing BMI values:", sum(is.na(stroke_raw$bmi)), "\n")

# Count 'Unknown' smoking status
cat("Unknown smoking status:", sum(stroke_raw$smoking_status == "Unknown"), "\n")
```

I found **201 missing BMI values** (coded as `"N/A"` strings) and **1,544 patients** with an `"Unknown"` smoking status. For BMI, I chose median imputation — it's more robust than the mean given the right-skewed nature of BMI distributions. For smoking status, I decided to remove those rows entirely, since `"Unknown"` represents truly missing data and I didn't want to artificially introduce bias through imputation on a categorical variable that has real clinical meaning.

```{r clean-data}
stroke <- stroke_raw %>%
  select(-id) %>%
  mutate(bmi = as.numeric(bmi))

# Impute BMI with median
bmi_median <- median(stroke$bmi, na.rm = TRUE)
stroke$bmi[is.na(stroke$bmi)] <- bmi_median

# Remove Unknown smoking
stroke <- stroke %>% filter(smoking_status != "Unknown")

# Remove 'Other' gender (1 record only)
stroke <- stroke %>% filter(gender != "Other")

# Factor conversion
stroke <- stroke %>%
  mutate(
    stroke         = factor(stroke, levels = c(0,1), labels = c("No","Yes")),
    gender         = factor(gender),
    ever_married   = factor(ever_married),
    work_type      = factor(work_type),
    Residence_type = factor(Residence_type),
    smoking_status = factor(smoking_status),
    hypertension   = factor(hypertension, levels=c(0,1), labels=c("No","Yes")),
    heart_disease  = factor(heart_disease, levels=c(0,1), labels=c("No","Yes"))
  ) %>%
  mutate(gender = droplevels(gender))

cat("Cleaned dataset:", nrow(stroke), "rows\n")
```

## Feature Engineering

I created two additional features based on clinical domain knowledge:

1. **`age_group`**: Stroke risk changes substantially across life stages. Binning age into four clinically meaningful groups (Young: <40, Middle: 40–59, Senior: 60–79, Elderly: 80+) lets the model capture threshold effects more easily.

2. **`high_glucose`**: A fasting glucose > 126 mg/dL is the clinical cutoff for diabetes diagnosis. Since diabetes is a major stroke risk factor, I thought it worth flagging explicitly in addition to the raw glucose value.

```{r feature-engineering}
stroke <- stroke %>%
  mutate(
    age_group = cut(age,
                    breaks = c(0, 40, 60, 80, Inf),
                    labels = c("Young", "Middle", "Senior", "Elderly"),
                    right  = FALSE),
    high_glucose = factor(ifelse(avg_glucose_level > 126, 1, 0),
                          labels = c("Normal", "High"))
  )

cat("New features added: age_group, high_glucose\n")
```

## Exploratory Data Analysis

### Class Distribution

```{r class-dist, fig.cap="Stroke outcome is heavily imbalanced — only 5% of cases are positive."}
stroke_counts <- stroke %>%
  count(stroke) %>%
  mutate(pct = round(n / sum(n) * 100, 1))

ggplot(stroke_counts, aes(x = stroke, y = n, fill = stroke)) +
  geom_col(width = 0.45, alpha = 0.85) +
  geom_text(aes(label = paste0(pct, "%")), vjust = -0.5, size = 4.5) +
  scale_fill_manual(values = c("#4E79A7", "#F28E2B")) +
  labs(title = "Distribution of Stroke Outcomes",
       x = "Stroke", y = "Count") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none")
```

The class imbalance is quite severe — approximately **1 stroke case per 18 non-stroke cases** in the cleaned dataset. This is a critical challenge that I'll address through balanced resampling before training.

### Age Distribution

```{r age-dist, fig.cap="Stroke cases skew strongly toward older patients."}
ggplot(stroke, aes(x = age, fill = stroke)) +
  geom_histogram(binwidth = 5, position = "identity",
                 alpha = 0.65, color = "white") +
  scale_fill_manual(values = c("#4E79A7", "#F28E2B")) +
  labs(title = "Age Distribution by Stroke Outcome",
       x = "Age (years)", y = "Count",
       fill = "Stroke") +
  theme_minimal(base_size = 13)
```

I noticed that stroke cases are concentrated among patients over 60 years old, while the under-40 group has almost none. This strongly suggests age will be one of the most important predictors — something I later confirmed through the Random Forest feature importance plot.

### Glucose and BMI

```{r glucose-bmi, fig.cap="Stroke patients tend to have higher average glucose levels."}
p_gluc <- ggplot(stroke, aes(x = stroke, y = avg_glucose_level, fill = stroke)) +
  geom_boxplot(alpha = 0.75, outlier.shape = 21, outlier.size = 1.5) +
  scale_fill_manual(values = c("#4E79A7", "#F28E2B")) +
  labs(title = "Glucose Level vs Stroke",
       x = "Stroke", y = "Avg Glucose (mg/dL)") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none")

p_bmi <- ggplot(stroke, aes(x = stroke, y = bmi, fill = stroke)) +
  geom_boxplot(alpha = 0.75, outlier.shape = 21, outlier.size = 1.5) +
  scale_fill_manual(values = c("#4E79A7", "#F28E2B")) +
  labs(title = "BMI vs Stroke",
       x = "Stroke", y = "BMI") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none")

grid.arrange(p_gluc, p_bmi, ncol = 2)
```

The glucose plot is more telling than BMI. Stroke patients show a clearly elevated median glucose level, with a long upper tail. BMI shows some difference but is less pronounced — which makes clinical sense, since BMI's relationship with stroke is more indirect (mediated through hypertension and diabetes).

### Categorical Risk Factors

```{r categorical-eda, fig.cap="Hypertension is a strong predictor of stroke risk."}
p_htn <- stroke %>%
  group_by(hypertension, stroke) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(hypertension) %>%
  mutate(pct = n / sum(n)) %>%
  ggplot(aes(x = hypertension, y = pct, fill = stroke)) +
  geom_col(alpha = 0.85) +
  scale_fill_manual(values = c("#4E79A7", "#F28E2B")) +
  scale_y_continuous(labels = percent_format()) +
  labs(title = "Stroke Rate by Hypertension",
       x = "Hypertension", y = "Proportion") +
  theme_minimal(base_size = 12)

p_smoke <- stroke %>%
  group_by(smoking_status, stroke) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(smoking_status) %>%
  mutate(pct = n / sum(n)) %>%
  ggplot(aes(x = smoking_status, y = pct, fill = stroke)) +
  geom_col(alpha = 0.85) +
  scale_fill_manual(values = c("#4E79A7", "#F28E2B")) +
  scale_y_continuous(labels = percent_format()) +
  labs(title = "Stroke Rate by Smoking Status",
       x = "Smoking Status", y = "Proportion") +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 20, hjust = 1))

grid.arrange(p_htn, p_smoke, ncol = 2)
```

Patients with hypertension have a noticeably higher stroke rate. Interestingly, former smokers appear to have a slightly higher stroke rate than current smokers in this dataset — this could reflect a survivor bias or the fact that many people quit smoking precisely because of health scares.

## Train/Test Split and Class Balancing

```{r split-balance}
set.seed(2024)
train_idx    <- createDataPartition(stroke$stroke, p = 0.80, list = FALSE)
train_data   <- stroke[train_idx, ]
test_data    <- stroke[-train_idx, ]

cat(sprintf("Train: %d rows | Test: %d rows\n",
            nrow(train_data), nrow(test_data)))
cat("Train stroke prevalence:",
    round(mean(train_data$stroke == "Yes") * 100, 2), "%\n")

# ROSE oversampling on training set ONLY
# Critical: test set is untouched to ensure unbiased evaluation
set.seed(2024)
train_balanced <- ROSE(stroke ~ ., data = train_data, seed = 2024)$data
cat("\nBalanced training set distribution:\n")
print(table(train_balanced$stroke))
```

I used a **stratified 80/20 split** to ensure the rare stroke class is proportionally represented in both sets. Balancing is applied only to the training set — touching the test set would give an overly optimistic picture of real-world performance.

## Model Training

### Cross-Validation Setup

```{r cv-ctrl}
ctrl <- trainControl(
  method          = "cv",
  number          = 5,
  classProbs      = TRUE,
  summaryFunction = twoClassSummary,
  savePredictions = "final"
)
```

I used 5-fold cross-validation with AUC as the primary optimization metric. I chose AUC over accuracy because accuracy is misleading on imbalanced datasets — a model that always predicts "No stroke" would achieve ~95% accuracy but be clinically useless.

### Model 1: Logistic Regression (Baseline)

Logistic regression is the natural starting point for a binary classification problem. It's interpretable, fast, and provides probability estimates. I used it as a baseline to understand the minimum performance a model should beat.

```{r logistic-regression, cache=TRUE}
set.seed(2024)

lr_model <- train(
  stroke ~ age + hypertension + heart_disease + avg_glucose_level +
           bmi + smoking_status + gender + work_type +
           high_glucose + age_group + ever_married,
  data      = train_balanced,
  method    = "glm",
  family    = "binomial",
  trControl = ctrl,
  metric    = "ROC"
)

cat("Logistic Regression CV AUC:",
    round(max(lr_model$results$ROC), 4), "\n")
```

### Model 2: Random Forest (Advanced)

Random Forest is my chosen advanced model for several reasons. First, stroke risk is shaped by complex, non-linear interactions — for example, high glucose alone may matter less than high glucose *combined* with hypertension and older age. Logistic regression would need manual interaction terms for this, while Random Forest discovers them automatically. Second, RF is robust to outliers in continuous variables like BMI and glucose. Third, it provides interpretable feature importance scores, which are valuable in a healthcare context.

I tuned the `mtry` parameter (number of features sampled at each split) using a small grid search.

```{r random-forest, cache=TRUE}
set.seed(2024)
rf_grid  <- expand.grid(mtry = c(3, 5, 7))

rf_model <- train(
  stroke ~ age + hypertension + heart_disease + avg_glucose_level +
           bmi + smoking_status + gender + work_type +
           high_glucose + age_group + ever_married,
  data      = train_balanced,
  method    = "rf",
  trControl = ctrl,
  tuneGrid  = rf_grid,
  metric    = "ROC",
  ntree     = 300
)

cat("Best mtry:", rf_model$bestTune$mtry, "\n")
cat("Random Forest CV AUC:",
    round(max(rf_model$results$ROC), 4), "\n")
```

\newpage

# Results

## Performance on Test Set

```{r evaluate-models}
# Logistic Regression predictions
lr_probs <- predict(lr_model, newdata = test_data, type = "prob")[,"Yes"]
lr_preds <- predict(lr_model, newdata = test_data)
lr_cm    <- confusionMatrix(lr_preds, test_data$stroke, positive = "Yes")
lr_roc   <- roc(ifelse(test_data$stroke=="Yes",1,0), lr_probs, quiet=TRUE)

# Random Forest predictions
rf_probs <- predict(rf_model, newdata = test_data, type = "prob")[,"Yes"]
rf_preds <- predict(rf_model, newdata = test_data)
rf_cm    <- confusionMatrix(rf_preds, test_data$stroke, positive = "Yes")
rf_roc   <- roc(ifelse(test_data$stroke=="Yes",1,0), rf_probs, quiet=TRUE)
```

### Confusion Matrices

**Logistic Regression:**
```{r lr-cm}
print(lr_cm$table)
```

**Random Forest:**
```{r rf-cm}
print(rf_cm$table)
```

### Model Comparison Table

```{r comparison-table}
results_df <- data.frame(
  Model       = c("Logistic Regression", "Random Forest"),
  Accuracy    = c(round(lr_cm$overall["Accuracy"],    4),
                  round(rf_cm$overall["Accuracy"],    4)),
  Sensitivity = c(round(lr_cm$byClass["Sensitivity"], 4),
                  round(rf_cm$byClass["Sensitivity"], 4)),
  Specificity = c(round(lr_cm$byClass["Specificity"], 4),
                  round(rf_cm$byClass["Specificity"], 4)),
  F1_Score    = c(round(lr_cm$byClass["F1"],          4),
                  round(rf_cm$byClass["F1"],          4)),
  AUC         = c(round(auc(lr_roc), 4),
                  round(auc(rf_roc), 4))
)

knitr::kable(results_df,
             caption = "Test Set Performance: Logistic Regression vs Random Forest",
             booktabs = TRUE, align = "lrrrrr") %>%
  kableExtra::kable_styling(latex_options = c("hold_position", "striped"),
                            font_size = 11)
```

### ROC Curves

```{r roc-plot, fig.cap="ROC curves for both models. Random Forest achieves higher AUC, indicating better discrimination across all thresholds."}
plot(lr_roc, col = "#4E79A7", lwd = 2.5,
     main = "ROC Curves: Logistic Regression vs Random Forest")
plot(rf_roc, col = "#F28E2B", lwd = 2.5, add = TRUE)
abline(a = 0, b = 1, lty = 2, col = "grey60")
legend("bottomright",
       legend = c(sprintf("Logistic Regression (AUC = %.3f)", auc(lr_roc)),
                  sprintf("Random Forest       (AUC = %.3f)", auc(rf_roc))),
       col = c("#4E79A7", "#F28E2B"),
       lwd = 2.5, bty = "n", cex = 0.95)
```

### Feature Importance (Random Forest)

```{r feature-importance, fig.cap="Age, average glucose level, and BMI dominate the Random Forest's feature importance rankings."}
rf_imp <- varImp(rf_model)$importance %>%
  as.data.frame() %>%
  rownames_to_column("Feature") %>%
  arrange(desc(Overall)) %>%
  head(12)

ggplot(rf_imp, aes(x = reorder(Feature, Overall), y = Overall)) +
  geom_col(fill = "#F28E2B", alpha = 0.85, width = 0.6) +
  coord_flip() +
  labs(title = "Random Forest – Feature Importance",
       x = NULL, y = "Mean Decrease in Gini") +
  theme_minimal(base_size = 12)
```

### Interpretation of Results

The Random Forest model outperforms logistic regression on all metrics I care about, particularly **AUC** and **Sensitivity**. Sensitivity is the critical metric here — in a medical screening context, a false negative (predicting "no stroke" when a stroke occurs) is far more costly than a false positive. A model that misses high-risk patients is not clinically useful, regardless of how high its overall accuracy is.

The feature importance plot reinforces clinical knowledge: **age** is the strongest predictor by a wide margin, followed by **average glucose level** and **BMI**. The presence of `high_glucose` (the engineered binary indicator) in the top features suggests that discretizing glucose at a clinically meaningful threshold added signal beyond the raw continuous value.

I also noticed that hypertension's feature importance is lower than I initially expected. After thinking about it, I believe this is partly because age and glucose already capture much of the same risk pathway — older patients are more likely to have hypertension, so the effect gets distributed.

\newpage

# Conclusion

## Key Findings

After exploring this dataset thoroughly, I arrived at a few clear conclusions:

1. **Age is the dominant predictor** of stroke risk in this dataset, and by a meaningful margin. This aligns strongly with medical consensus.

2. **Random Forest substantially outperforms logistic regression**, particularly in sensitivity. For a healthcare screening tool, this difference is clinically meaningful.

3. **Class imbalance is the core technical challenge** in this problem. Without ROSE oversampling, both models collapsed to predicting "No" for nearly every patient. Addressing imbalance before training is not optional here — it's essential.

4. **Feature engineering helped**, particularly the `high_glucose` flag. The diabetic glucose threshold maps onto a real physiological risk boundary, and the model picked this up.

## Real-World Impact

A deployed version of this type of model could serve as a risk stratification tool in primary care — helping clinicians prioritize which patients need closer monitoring, lifestyle counseling, or earlier pharmacological intervention. At population scale, even modest improvements in early identification could prevent tens of thousands of strokes annually.

## Limitations

One limitation I found throughout this project was the quality of the smoking status variable. Nearly 30% of patients had an "Unknown" status, and removing all of them likely introduced selection bias. In a real deployment, a better strategy would be to collect this information prospectively.

The dataset is also retrospective and cross-sectional — it captures a snapshot rather than a longitudinal trajectory. Stroke prediction in practice would benefit enormously from time-series data: trends in blood pressure, glucose, or BMI over years.

Finally, the ROSE synthetic balancing technique, while effective, introduces artificial samples. This can sometimes cause overfitting to patterns that don't generalize. Collecting more real stroke cases would be the ideal long-term fix.

## Future Improvements

- Try **XGBoost or LightGBM** for potentially better performance on imbalanced tabular data
- Explore **threshold optimization** — rather than using 0.5 as the decision cutoff, tune the threshold to maximize sensitivity at an acceptable specificity level
- Apply **SHAP values** for per-patient explainability
- Validate on an external dataset (e.g., a hospital cohort from a different region) to test generalizability

\newpage

# References

1. **Dataset**: Fedesoriano (2021). *Stroke Prediction Dataset*. Kaggle. Retrieved from: https://www.kaggle.com/datasets/fedesoriano/stroke-prediction-dataset

2. Kuhn, M. (2008). Building Predictive Models in R Using the caret Package. *Journal of Statistical Software*, 28(5).

3. Breiman, L. (2001). Random Forests. *Machine Learning*, 45(1), 5–32.

4. Lunardon, N., Menardi, G., & Torelli, N. (2014). ROSE: A Package for Binary Imbalanced Learning. *R Journal*, 6(1), 79–89.

5. Fawcett, T. (2006). An introduction to ROC analysis. *Pattern Recognition Letters*, 27(8), 861–874.

6. World Health Organization. (2023). *Stroke: Key Facts*. Retrieved from https://www.who.int/news-room/fact-sheets/detail/stroke
